Leadership across cultures and contexts: an exploratory analysis

### 2020-06-24 leadership_across_cultures_contexts.Rmd
Minor edits.

<!-- WEIRD PEOPLE PROBLEMS -->
There is now a broad consensus on the importance of data from the full range of human cultural diversity. Since Henrich et al.'s [-@henrich_weirdest_2010] seminal WEIRD people paper, many social scientists have sought data from "non-WEIRD" populations. Dichotomizing populations, their psychologies, or social dynamics as WEIRD or non-WEIRD, however, is a mistake.^[A Google Scholar search for "non-WEIRD" produced over 1000 results.] As anthropology has conclusively demonstrated, there is enormous diversity across societies that does not remotely resemble a dichotomy.

Anthropology, on the other hand, has been criticized for essentializing diverse "Others" as part of its colonial history [@said1979; @abu2008]. In this regard, @abu2008 noted that the concept of culture operates much like its predecessor -- race.

@@

Our fourth goal was to explore lower dimensional representations of these data that might reveal informative "features." These included hierarchical cluster analysis, logistic PCA, and minimal spanning trees and k-nearest neighbors (mst-kNN). See the SI for additional details. Our fifth goal was to assess variation in our lower-dimensional "features" by group context, subsistence strategy, continental region, and leader sex, using logistic mixed effects models fit with the lme4 package. Our sixth goal was to evaluate the features in light of evolutionary theories of leadership. Guided by empirically derived features, we aimed to construct "elements" of leadership that corresponded to elements of influential theories, examining intersections of these elements across all text records.

@@

Words that were strong predictors of leaders with coercive authority included, unsurprisingly, *power* and *authority*, as well as *territory*, *chief*, and *control*; the word *leadership* was the only strong negative predictor. Words that predicted no coercive authority tended to be those that indicated either hunter-gatherers, such as "band" and "Mbuti" (Congo Basin foragers), or kinship and residential-level contexts, such as "lineage", "house", "clan", and "village". See Figures \@ref(fig:elasticcoercive)A,C.

@@

The leader functions cluster analysis (Figure \@ref(fig:cluster)B) also identified two top-level clusters; the larger of which we identified as a management cluster and the smaller of which we identified as a *Prosociality* cluster. Within the management cluster, we identified two moderately supported subclusters: *Strategize* and *Organize* subclusters.

@@

The leadership features we derived by clustering leadership dimensions were represented across most cultures. Leaders performed three broad functions: providing prosocial services, strategizing, and organizing collective actions. Regarding characteristic traits, leaders possessed some combination of individual capacities, many of which are prosocial, conformed to cultural norms, and receivied various benefits (Figure \@ref(fig:featuressupportplot)).

These potentially universal dimensions of leadership correspond to several evolutionary theories that have drawn on select ethnographic cases [@garfield_review_2019]. They partially correspond to the prestige model [@henrich_big_2015; @henrich_evolution_2001], which emphasizes knowledge, skill and generosity, but they do not include a key feature of that model: emulation of prestigious leaders. We labeled one of our leader quality features "cultural conformity", however, in which leaders are culturally conservative and exemplify desirable traits such as honesty, humility, and "proper behavior" (Figure \@ref(fig:cluster)). This feature corresponds in part to leaders who serve as models for their followers [see also @garfield_investigating_2020].

@@

Evidence of coercive authority is common in the ethnographic record of leadership and does not vary by any of our four factors, yet did not quite make the 60% cutoff for our candidate universals. This dimension features in several evolutionary models of leadership involving dominance hierarchies [@barkow_darwin_1989; @tiger_imperial_1971; @henrich_evolution_2001; @cheng_dominance_2019; @van_vugt_dual_2019] and the role of punishment in promoting cooperation [e.g., @marlowe_more_2008; @henrich_markets_2010; @henrich_costly_2006; @boyd_punishment_1992]. Indeed, we found that coercive authority among leaders was associated with enforcing punishment, aggressiveness, being feared, and also with controlling economics, which evokes "resource holding power", a concept from the behavioral ecology of dominance hierarchies (Figure \@ref(fig:elasticcoercive)). The authority of Trobriand chiefs is illustrative [@oj29-007, pp. 64-65]:

@@

These four factors were confounded. Evidence for leadership in residential subgroups and kin groups was relatively more common in hunter-gatherer and pastoralist societies, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013]. Evidence for supracommunity political leadership was more common in horticultural and agricultural societies, which tend to rely heavily on cultivated land and maintain fixed, defensible territories (Figures \@ref(fig:grouptype) and S\@ref(fig:regionsubsistence)). These patterns support perspectives suggesting residential mobility and subsistence intensification are associated with increasing complexity of sociopolitical organization [@johnson_evolution_1987; @steward_basin-plateau_1938; @powers_evolutionary_2014]. Perhaps as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups diminishes.

@@

When ethnographic texts described female leaders exclusively, they were relatively more likely to be described as married to a high-quality spouse in a polygynous marriage, and the recipients of various mating benefits (Figure S\@ref(fig:heatmapsex)). Female leaders also had higher levels of evidence for the *Social, reproductive, and material success* "feature" (Figure S\@ref(fig:heatmapfeaturesex)). Evidence for female leaders was also more likely to be found in kin and residential subgroups. This suite of characteristics supports discussions of high status first-wives who gain and maintain social influence across the lifespan by leveraging their extended kin and social networks [@goodale_tiwi_1971; @brown_her_1985; @yanca_female_2004]. @garfield_investigating_2020 similarly found evidence of positive assortative mating of high status leaders in a small-scale society with relative gender-egalitarianism. Our results are also consistent with theoretical arguments emphasizing women as leaders of families [e.g., @garfield_evolutionary_2019; @hagen_leadership_2019].

@@

Leaders across cultures rely a range of individual capacities, including cognitive, supernatural, material, social, and physical endowments, to organize group members, implement strategic actions, provide prosocial services to the group, and impose costs, all while conforming to cultural norms. Currently, no single theoretical perspective has yet captured the ethnographic reality of human leadership.

>>>>

<!-- WEIRD PEOPLE PROBLEMS -->
There is now a broad consensus on the importance of data from the full range of human cultural diversity. Since Henrich et al.'s [-@henrich_weirdest_2010] seminal WEIRD people paper, many social scientists have sought data from "non-WEIRD" populations. Dichotomizing populations, their psychologies, or social dynamics as WEIRD or non-WEIRD, however, is a mistake.^[A Google Scholar search for "non-WEIRD" produced over 1000 results.] As anthropology has conclusively demonstrated, there is enormous diversity across societies that does not remotely resemble a dichotomy. Conversely, anthropology has been criticized for essentializing diverse "Others" as part of its colonial history [@said1979; @abu2008]. In this regard, @abu2008 noted that the concept of culture operates much like its predecessor -- race.

@@

Our fourth goal was to explore lower dimensional representations of these data that might reveal informative "features." These included hierarchical cluster analysis, logistic PCA, and minimal spanning trees and k-nearest neighbors (mst-kNN). See the SI for additional details. Our fifth goal was to assess variation in our lower-dimensional features by group context, subsistence strategy, continental region, and leader sex, using logistic mixed effects models fit with the lme4 package. Our sixth goal was to evaluate the features in light of evolutionary theories of leadership. Guided by empirically derived features, we aimed to construct "elements" of leadership that corresponded to elements of influential theories, examining intersections of these elements across all text records.

@@

Words that were strong predictors of leaders with coercive authority included, unsurprisingly, *power* and *authority*, as well as *territory*, *chief*, and *control*; the word *leadership* was the only strong negative predictor. Words that predicted no coercive authority tended to be those that indicated either hunter-gatherers, such as *band* and *Mbuti* (Congo Basin foragers), or kinship and residential-level contexts, such as *lineage*, *house*, *clan*, and *village*. See Figures \@ref(fig:elasticcoercive)A,C.

@@

The leader functions cluster analysis (Figure \@ref(fig:cluster)B) also identified two top-level clusters; the larger of which we identified as a management cluster and the smaller of which we identified as a *Prosociality* cluster. Within the management cluster, we identified two moderately supported subclusters: *Strategize* and *Organize*.

@@

The leadership features we derived by clustering leadership dimensions were represented across most cultures. Leaders performed three broad functions: providing prosocial services, strategizing, and organizing collective actions. Regarding characteristic traits, leaders possessed some combination of individual competencies, many of which are prosocial, conformed to cultural norms, and receivied various benefits (Figure \@ref(fig:featuressupportplot)).

These potentially universal dimensions of leadership correspond to several evolutionary theories that have drawn on select ethnographic cases [@garfield_review_2019]. They partially correspond to the prestige model [@henrich_big_2015; @henrich_evolution_2001], which emphasizes knowledge, skill and generosity, but they do not include a key feature of that model: emulation of prestigious leaders. We labeled one of our leader quality features "cultural conformity", however, in which leaders are culturally conservative and exemplify desirable traits such as honesty, humility, and "proper behavior" (Figure \@ref(fig:cluster)). This feature corresponds in part to leaders who serve as models for followers [see also @garfield_investigating_2020].

@@

Evidence of coercive authority is common in the ethnographic record of leadership and does not vary by any of our four factors, yet did not quite make the 60% cutoff for our candidate universals. This dimension features in several evolutionary models of leadership involving dominance hierarchies [@barkow_darwin_1989; @tiger_imperial_1971; @henrich_evolution_2001; @cheng_dominance_2019; @van_vugt_dual_2019] and the role of punishment in promoting cooperation [e.g., @marlowe_more_2008; @henrich_markets_2010; @henrich_costly_2006; @boyd_punishment_1992]. Indeed, we found that coercive authority among leaders was associated with enforcing punishment, aggressiveness, being feared, and also with controlling economics, which evokes "resource holding power," a concept from the behavioral ecology of dominance hierarchies (Figure \@ref(fig:elasticcoercive)). The authority of Trobriand chiefs is illustrative [@oj29-007, pp. 64-65]:

@@

These four factors were confounded. Evidence for leadership in residential subgroups and kin groups was relatively more common in hunter-gatherer and pastoralist societies, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013]. Evidence for supracommunity political leadership was more common in horticultural and agricultural societies, which tend to rely heavily on cultivated land and maintain fixed, defensible territories (Figures \@ref(fig:grouptype) and S\@ref(fig:regionsubsistence)). These patterns support perspectives suggesting restricted residential mobility and subsistence intensification are associated with increasing complexity of sociopolitical organization [@johnson_evolution_1987; @steward_basin-plateau_1938; @powers_evolutionary_2014]. Perhaps as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups diminishes.

@@

When ethnographic texts described female leaders exclusively, they were relatively more likely to be described as married to a high-quality spouse in a polygynous marriage, and the recipients of various mating benefits (Figure S\@ref(fig:heatmapsex)). Female leaders also had higher levels of evidence for the *Social, reproductive, and material success* feature (Figure S\@ref(fig:heatmapfeaturesex)). Evidence for female leaders was also more likely to be found in kin and residential subgroups. This suite of characteristics supports discussions of high status first-wives who gain and maintain social influence across the lifespan by leveraging their extended kin and social networks [@goodale_tiwi_1971; @brown_her_1985; @yanca_female_2004]. @garfield_investigating_2020 similarly found evidence of positive assortative mating of high status leaders in a small-scale society with relative gender-egalitarianism. Our results are also consistent with theoretical arguments emphasizing women as leaders of families [e.g., @garfield_evolutionary_2019; @hagen_leadership_2019].

@@

Leaders across cultures rely a range of individual competencies, including cognitive, supernatural, material, social, and physical endowments, to organize group members, implement strategic actions, provide prosocial services to the group, and impose costs, all while conforming to cultural norms. Currently, no single theoretical perspective has yet captured the ethnographic reality of human leadership.
